Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-04-30 thesis/draft.tex
temp storage

Not surprisingly, employing the GCT model with 2 substages for both generating and fitting incidence data also resulted in a close match, as indicated by an RMSLE of 2.34\% (Figure \ref{sametypeFittingG}.A), and parameters post-fitting were close to the exact values (Figure \ref{sametypeFittingG}.B). Unlike the case with the LCT model, when the structure of the fitting model was altered from 2 substages to 6, the model still exhibited a close match with the original in terms of both shape (Figure \ref{sametypeFittingG}.C) and parameters (Figure \ref{sametypeFittingG}.D). This indicates that the GCT model's fitting performance is minimally impacted by changes in the model structure.

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0) {\includegraphics[width= \textwidth]{figures/sametypeFittingG.Rout.pdf}};
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % Add labels at desired positions
            \node at (0.01,0.95) {\textbf{(A)}};
            \node at (0.55,0.95) {\textbf{(B)}};
            \node at (0.01,0.45) {\textbf{(C)}};
            \node at (0.55,0.45) {\textbf{(D)}};
        \end{scope}
    \end{tikzpicture}
    \caption{\textbf{Model fitting and parameter estimation in GCT model.} \\ Data generated from the GCT model featured 12 substages, with $\beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$ and $\kappa=\frac{2}{9}$. Remaining details are as described in Figure \ref{sametypeFittingL}.}
    \label{sametypeFittingG}
\end{figure}

%%%%%%%%
% Here
%%%%%%%%
To more clearly illustrate the performance differences between the LCT and GCT models, fittings with various model structures were conducted to assess changes in percentage error. The percentage error from the fitting result, where the model generating the incidence data and the model for fitting are exactly the same, has been used as the baseline (red line). As shown in Figure X, when using the same input data, varying the fitting model shape causes the LCT model to experience a surge in RMSLE values, while the GCT model maintains its error rate close to the baseline. This result, once again, confirms that the structure of the GCT model merely affects the fitting results, indicating that changes in structure to optimize model performance are no longer necessary.

To inllstriate more clear about the performance difference between LCT and GCT, fittings using different model structure have been done to see how the error would change. As shown in Figure x, when treating the percentage error from the the fitting result where the model to generate the incidence data and the model for fitting are exactly the same, varying the fitting model shape by using a different number of substages from the model used to generate the data causes the LCT model to experience a surge in RMSLE values, while the GCT model maintains its error rate close to the baseline. This result, once again, confirmed that the structure of the GCT model merely affect the model fitting result, showing that changing in structure in order to find the optimized performance of the model is no longer necessary.

@@

\subsection{Parameter Estimation}
\label{parm est}
In this section, we discuss the parameter estimation results obtained through model fitting. The true incidence data were generated using the GCT model with 5 substages, setting the $D$ to 10 and $\kappa$ to $\frac{3}{10}$. Random noise was added to the true incidence data using a negative binomial distribution, resulting in four different input datasets (a number we specifically set for this analysis). The model was fitted using both the GCT and LCT frameworks with 4 to 12 substages using the 4 sets of data. As depicted in Figure \ref{parEst}, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the estimated parameter values derived from the LCT model, using different input data, tend to cluster together but on average vary across different substage numbers. These estimates could deviate substantially from the accurate parameter if an inappropriate structure is chosen. On the other hand, although the parameter estimation using the GCT model is not perfect—with kappa and beta being overestimated and D underestimated—the estimated values are less fluctuating across substage numbers, closer to the target values, and generally include the target values within the 95\% confidence interval. The discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low root mean squared logarithmic error (RMSLE).

>>>>

Not surprisingly, employing the GCT model with 12 substages for both generating and fitting incidence data also resulted in a close match, as indicated by an RMSLE of 2.34\% (Figure \ref{sametypeFittingG}.A), and parameters post-fitting were close to the exact values (Figure \ref{sametypeFittingG}.B). Unlike the case with the LCT model, when the structure of the fitting model was altered from 12 substages to 6, the model still exhibited a close match with the original in terms of both shape (Figure \ref{sametypeFittingG}.C) and parameters (Figure \ref{sametypeFittingG}.D). This indicates that the GCT model's fitting performance is minimally impacted by changes in the model structure.

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0) {\includegraphics[width= \textwidth]{figures/sametypeFittingG.Rout.pdf}};
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % Add labels at desired positions
            \node at (0.01,0.95) {\textbf{(A)}};
            \node at (0.55,0.95) {\textbf{(B)}};
            \node at (0.01,0.45) {\textbf{(C)}};
            \node at (0.55,0.45) {\textbf{(D)}};
        \end{scope}
    \end{tikzpicture}
    \caption{\textbf{Model fitting and parameter estimation in GCT model.} \\ Data generated from the GCT model featured 2 substages, with kappa= $\beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$ and $\kappa=\frac{2}{9}$. Remaining details are as described in Figure \ref{sametypeFittingL}.}
    \label{sametypeFittingG}
\end{figure}

%%%%%%%%
% Here
%%%%%%%%
To more clearly illustrate the performance differences between the LCT and GCT models, fittings with various model structures were conducted to assess changes in percentage error. The percentage error from the fitting result, where the model generating the incidence data and the model for fitting are exactly the same, has been used as the baseline (red line). As shown in Figure \ref{rmsle}, when using the same input data, varying the fitting model shape causes the LCT model to experience a surge in RMSLE values, while the GCT model maintains its error rate close to the baseline. This result, once again, confirms that the structure of the GCT model merely affects the fitting results, indicating that changes in structure to optimize model performance are no longer necessary.

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{figures/rmsle.Rout.pdf}
    \caption{\textbf{Percentage error with different fitting model structures.} \\ The LCT fitting data were generated by simulating a two-substage LCT transmission model with $D=10, \beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$. Similarly, GCT fitting data came from simulating the GCT model with identical properties but $\kappa=\frac{5}{9}$. The baseline (red) represents the error rate when the fitting and data generation models have the same structure. The blue circle indicates LCT errors, and the orange rhombus represents GCT errors.}
    \label{rmsle}
\end{figure}

@@

\subsection{Parameter Estimation}
\label{parm est}
In this section, we discuss the parameter estimation results obtained through model fitting. The true incidence data were generated using the GCT model with 5 substages, setting the $D$ to 10 and $\kappa$ to $\frac{3}{10}$. Random noise was added to the true incidence data using a negative binomial distribution, resulting in four different input datasets (a number we specifically set for this analysis). The model was fitted using both the GCT and LCT frameworks with 4 to 12 substages using the 4 sets of data. As depicted in Figure \ref{parEst}, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the estimated parameter values derived from the LCT model, using different input data, tend to cluster together but on average vary across different substage numbers. These estimates could deviate substantially from the accurate parameter if an inappropriate structure is chosen. On the other hand, although the parameter estimation using the GCT model is not perfect—with kappa and beta being underestimated and D overestimated—the estimated values are less fluctuating across substage numbers, closer to the target values, and generally include the target values within the 95\% confidence interval. The discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low root mean squared logarithmic error (RMSLE).
